---
title: "LMM Analysis Fixation Duration"
author: "Schwetlick Lisa, Backhaus Daniel"
date: "July 2021"
last change: "minor rmd style"
output:
  slidy_presentation:
  css: font.css
beamer_presentation: default
ioslides_presentation: default
header-includes: \usepackage{amsmath}
editor_options:
chunk_output_type: console
---
  
  
```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
#setwd("/Users/Daniel/CloudStation/Arbeit/03_Allgemeine Psychologie/2020/Lisa_Memory/SUBSET_SMLP/")

set.seed(123)

library(MASS)
library(lme4)
library(tidyverse)

ginv2 <- function(x) # define a function to make the output nicer
  MASS::fractions(provideDimnames(MASS::ginv(x),base=dimnames(x)[2:1]))

theme_set(theme_bw(base_size=18))
```

# Load data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Filename<-"smlp_mem_data.RData"
load(Filename)#d
```

# Visualization of fixation duration

```{r,echo=FALSE, message=FALSE, warning=FALSE}
fd<-d%>%
  group_by(Experiment, FPT, imtype, Subject, FSP)%>%
  summarise(mfd=mean(fixdur, na.rm=T), N=n())%>%
  group_by(Experiment, Subject)%>%
  mutate(mSubject=mean(mfd, na.rm = T))%>%
  mutate(M=nrow(unique(data.frame(Experiment,FPT,imtype,FSP))) )%>%
  ungroup(.)%>%
  mutate(mall = mean(mfd, na.rm=T))%>%
  mutate(mcos=mfd-mSubject+mall)%>%
  group_by(Experiment, imtype, FPT, FSP)%>%
  summarise(m=mean(mfd, na.rm=T),
            N=n(),
            sd=(sd(mcos, na.rm=T)*sqrt((unique(M)/(unique(M)-1)))), # von daniel
            se=sd/sqrt(N-1)) %>% 
  mutate(tog= paste(Experiment,FSP))
```


```{r, fig.width=14, fig.height=5}
(fd_plot<- ggplot(data=fd, aes(x=factor(FPT), y=m, color=Experiment, group=tog, linetype=FSP))+
  facet_grid( ~ imtype, scales = "free_x")+
  geom_line(size=1.5)+
  geom_point(size=3, aes(shape=Experiment))+
  geom_errorbar(aes(ymin=m-qt(.975,N-1)*se, ymax=m+qt(.975,N-1)*se), width=.2, size=.75)+ # confidence intervals
  ggtitle("Fixation Duration")+
  xlab("Nth Viewingn (FPT)")+
  ylab("Mean Fixation Duration [ms]")+
  labs(linetype = "Similar Possible (FSP)", color="Experiment"))
```

# Research question

This experiment looks at eye movement characteristics and the effect of seeing the same image twice.
We call this "presentation type" (first vs. second viewing)
Additionally we manipulated whether the subjects think  there is a possibility similar images being shown.
We call this factor "similar possible".
In the data we will look at do not actually contain any similar images. The manipulation only refers to the
instruction and previous or later tasks.

 
columns:

* Experiment = we conducted three separate experiments with different subjects. We used same images in all three experiments.The only difference is in the factor similar possible(FSP).
* fixdur = fixation duration our dependent variable
* Img = image number
* trial = trial number
* Subject = subject number
* nthpres = ordinal presentation number of an image: "1", "2"
* imtype = imagetype, is an image: "new", "identical to a former new image" or "similar to a former new image" - this last type we exclude for this analysis
* FPT = factor presentation type: "new", "identical" 
* FSP = factor similar possible: did the subjects think that similar images could appear? "FALSE" "TRUE

Note: the factor similar possible (FSP) is a between subject factor for experiment 3 and 4 
      BUT a within-subject factor in experiment 5


# Select relevant data and setup dependent variable

```{r}
d<-d  %>% 
  filter(imtype != "similar") %>%  #not of interest for this analysis
  select(-c(imtype)) %>% 
  mutate(FPT = factor(FPT)) %>% 
  mutate(DV=fixdur, DV_org=DV, DV_log=log(DV), DV_reci=1/DV) 

# have a look
str(d)
```


# Create custom contrasts

```{r}
# Create custom contrasts FPT (factor presentation type)
t(Xicost <- fractions(rbind(Hyp1=c(X1= -1 ,  X2= 1))))

(Xcost <- ginv2(Xicost))
colnames(Xcost) <- c("FNewIdentical")
contrasts(d$FPT) <- Xcost


# Create custom contrasts FSP (factor similar possible)
t(Xicost <- fractions(rbind(Hyp1=c(X1= -1 ,  X2= +1))))   # could be changed to treatment if this is of interest


(Xcost <- ginv2(Xicost))
colnames(Xcost) <- c("FSimilarPossible")
contrasts(d$FSP) <- Xcost

mat_myC <- model.matrix(~ FPT + FSP, d)
```

## Check collinearity

```{r}
#find perfectly multicollinear variables
constant<-rep(1,nrow(mat_myC))
tmp<-lm(constant ~ ., data=as.data.frame(mat_myC))
(collinear<-tmp$coefficients[is.na(tmp$coefficients)]) # ignore the intercept here
# no columns are collinear


# add modelmatrix to dataframe
ncol(mat_myC)
start<-ncol(d)+1
end<-ncol(d)+ncol(mat_myC)
d[,start:end]<-mat_myC

names(d)[start:end] <- c("Intercept","FNewIdentical","FSimilarPossible" )
```

## Have a look

```{r}
head(d)
summary(d)
glimpse(d)
```


# Fit minimal model

```{r}
lmm.min <- lmer(DV ~ 1+ FPT + FSP  + (1| Subject) + (1| Img), data=d, REML=FALSE)

summary(rePCA(lmm.min))
summary(lmm.min)
```


# Question

We have three experiments. In experiment 3 similar images were possible. In experiment 4 similar images were not possible.
In experiment 5 we used "similar possible" as a within-subject variable in subsequent blocks. 
So we have the same factor in all three experiments, but it is a within-subject factor only for experiment 5 and a between subject factor for 3 and 4.

How can we implement this in our varying effect (random) part of the model formula?

```{r}
lmm.1 <- lmer(DV ~ 1+ FPT + FSP  + (1+ FPT + FSP| Subject) + (1+ FPT + FSP| Img), data=d, REML=FALSE)
```